Generalized Bradley-Terry models

### 2025-02-18 inst/jss_hyper3_real.Rnw
cheering (tikz diagram is bad atm)

% -*- mode: noweb; noweb-default-code-mode: R-mode; -*-
\documentclass[article]{jss}
\usepackage{booktabs}
\usepackage{units}
\usepackage{amsmath}
\usepackage{amssymb}

@@

\begin{figure}[htbp]
  \begin{center}
<<makecontour, cache=TRUE>>=
n <- 2
lambda <- seq(from=1, to=3, len=n)
D <- seq(from=0.1, to=0.6, len=n)
V <- as.matrix(expand.grid(lambda, D))
LL <- apply(V, 1, f)
LL <- pmax(-20, LL - max(LL))
contour(lambda, D, matrix(LL, n, n),levels = -2*(0:9), xlab=expression(lambda),ylab=expression(D))
points(maxv[1], maxv[2], pch=16, col='red')
@
\caption{Contours of support\label{contourlambdaD} evaluated over
  a grid.  Maximum likelihood estimate for $(\lambda, D)$ shown as
  a red dot at the evaluate}
  \end{center}
\end{figure}

@@

>>>>

% -*- mode: noweb; noweb-default-code-mode: R-mode; -*-
\documentclass[article]{jss}
\usepackage{booktabs}
\usepackage{units}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{tikz}

@@

\begin{figure}[htbp]
  \begin{center}
<<makecontour, cache=TRUE>>=
n <- 20
lambda <- seq(from=1, to=3, len=n)
D <- seq(from=0.1, to=0.6, len=n)
V <- as.matrix(expand.grid(lambda, D))
LL <- apply(V, 1, f)
LL <- pmax(-20, LL - max(LL))
contour(lambda, D, matrix(LL, n, n),levels = -2*(0:9), xlab=expression(lambda),ylab=expression(D))
points(maxv[1], maxv[2], pch=16, col='red')
@
\caption{Contours of support\label{contourlambdaD} evaluated over
  a grid.  Maximum likelihood estimate for $(\lambda, D)$ shown as
  a red dot at the evaluate}
  \end{center}
\end{figure}

@@

\section{Non-independence in successive arrivals}

In the following, I use the ``race'' metaphor: five runners take part
in a race and arrive in order: $a\succ b\succ c\succ d\succ e$.
Consider the following Plackett-Luce likelihood function for this
observation:

\begin{equation}
\frac{p_a}{p_a + p_b + p_c + p_d + p_e}\cdot
\frac{p_b}{      p_b + p_c + p_d + p_e}\cdot
\frac{p_c}{            p_c + p_d + p_e}\cdot
\frac{p_d}{                  p_d + p_e}\cdot
\frac{p_e}{                        p_e}
\end{equation}

Note the use of multiplication between the terms, indicating
conditional independence.  We may use the multiplicative
generalization of Bradley-Terry strengths to introduce an element of
non-independence between the terms.  There are many ways to do this,
but one simple case would be to define equivalence classes of the
competitors with each equivalence class comprising mutually supporting
runners.  Equivalence classes of size one correspond to unsupported
runners.  The metaphor would be that a runner who has finished the
race is able to support other members of his equivalence class by
cheering his teammates, boosting their performance.  The \pkg{hyper2}
package includes function \code{cheering()} which implements this
functionality.

As an example, consider figure~\ref{tikzabcde}.  This shows a partial
probability tree diagram for some of the $5!=120$ possible order
statistics.  The standard Plackett-Luce likelihoods have been modified
to account for two groups of mutually supporting runners:
$\left\lbrace a,b,c\right\rbrace$ with support term $\lambda$, and
$\left\lbrace d,e\right\rbrace$ with support term $\mu$.  From START,
the first runner to cross the finishing line has standard
Plackett-Luce probability.  Taking the top path as an example, we see
that the likelihood function for $a\succ b\succ c\succ d\succ e$ would
be

\begin{figure}[htbp]
  \begin{center}
\usetikzlibrary{arrows}
\usetikzlibrary{patterns}
\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=1cm,y=1cm]
\fill (0,0) circle[radius=2pt]; % root; paths abcde
\draw (0,0) -- (2, 3) -- (4.8, 3);
\draw (0,0) -- (2, 2) -- (4.8, 2);
\draw (0,0) -- (2, 1) -- (4.8, 1);
\draw (0,0) -- (2, 0) -- (4.8, 0);
\draw (0,0) -- (2,-1) -- (4.8,-1);
\node at (0,2.5) (eq1) {$\underbrace{\left\lbrace a,b,c\right\rbrace}_{\lambda}\underbrace{\left\lbrace d,e\right\rbrace}_{\mu}$};

\node at (4.8, 3) {$a$};
\node at (4.8, 2) {$b$};
\node at (4.8, 1) {$c$};
\node at (4.8, 0) {$d$};
\node at (4.8,-1) {$e$};

\node at (3.4, 3.4)  {$\frac{p_a}{p_a+p_b+p_c+p_d+p_e}$};
\node at (3.4, 2.4)  {$\frac{p_b}{p_a+p_b+p_c+p_d+p_e}$};
\node at (3.4, 1.4)  {$\frac{p_c}{p_a+p_b+p_c+p_d+p_e}$};
\node at (3.4, 0.4)  {$\frac{p_d}{p_a+p_b+p_c+p_d+p_e}$};
\node at (3.4,-0.6)  {$\frac{p_e}{p_a+p_b+p_c+p_d+p_e}$};

\fill  (4.8, 3) circle[radius=2pt];  % a finishes; paths bcde
\fill  (4.8, 2) circle[radius=2pt];  % terminal node
\fill  (4.8, 1) circle[radius=2pt];  % terminal node
\fill  (4.8, 0) circle[radius=2pt];  % d finishes; paths abce
\fill  (4.8,-1) circle[radius=2pt];  % terminal node

\draw (4.8,3) -- (5.8,5);
\draw (4.8,3) -- (5.8,4);
\draw (4.8,3) -- (5.8,3);
\draw (4.8,3) -- (5.8,2);

\node at (4.5,4.6) {$b$};
\node at (4.5,3.7) {$c$};
\node at (4.5,3.2) {$d$};
\node at (4.5,2.7) {$e$};

\newcommand{\la}[1]{\lambda{#1}}
\newcommand{\ld}[1]{\mu    {#1}}
\node at (6, 5)  {$\frac{\la{p_b}}{\la{p_b}+\la{p_c}+p_d+p_e}$};
\node at (6, 4)  {$\frac{\la{p_c}}{\la{p_b}+\la{p_c}+p_d+p_e}$};
\node at (6, 3)  {$\frac{p_d     }{\la{p_b}+\la{p_c}+p_d+p_e}$};
\node at (6, 2)  {$\frac{p_e     }{\la{p_b}+\la{p_c}+p_d+p_e}$};

\fill (7.4, 5) circle[radius=2pt];  % ab finishes; pahts cde
\draw (7.4,5) -- (8.4,6);
\draw (7.4,5) -- (8.4,5);
\draw (7.4,5) -- (8.4,4);

\node at (7.9,5.7) {$p_c$};
\node at (7.9,5.2) {$p_d$};
\node at (7.9,4.7) {$p_e$};
\node at (9.3, 6) {$\frac{\la{p_c}}{\la{p_c}+p_d+p_e}$};
\node at (9.3, 5) {$\frac{p_d  }{\la{p_c}+p_d+p_e}$};
\node at (9.3, 4) {$\frac{p_e  }{\la{p_c}+p_d+p_e}$};

\fill (10.2, 6) circle[radius=2pt];  % abc finishes; paths de
\draw (10.2, 6) -- (11.2,7);
\node at (10.7,6.8) {$p_d$};

\node at (11.8, 7) {$\frac{p_d}{p_d+p_e}$};

\draw (10.2, 6) -- (11.2,6.2);
\node at (11.8, 6.2) {$\frac{p_e}{p_d+p_e}$};
\node at (10.7,6.3) {$p_e$};

\fill (10.2, 5) circle[radius=2pt];
\draw (10.2, 5) -- (11.2,5.3);
\node at (11.8, 5.3) {$\frac{\la{p_c}}{\la{p_c}+\ld{p_e}}$};
\node at (10.7,5.3) {$p_c$};

\draw (10.2, 5) -- (11.2,4.7);
\node at (11.8, 4.7) {$\frac{\ld{p_e}}{\la{p_c}+\ld{p_e}}$};
\node at (10.7,4.7) {$p_e$};

\fill (10.2, 4) circle[radius=2pt];
\draw (10.2, 4) -- (11.2,3.9);
\node at (11.8, 3.9) {$\frac{\la{p_c}}{\la{p_c}+\ld{p_e}}$};
\node at (10.7,4.1) {$p_c$};

\draw (10.2, 4) -- (11.2,3.1);
\node at (11.8, 3.1) {$\frac{\ld{p_e}}{\la{p_c}+\ld{p_e}}$};
\node at (10.7,3.7) {$p_e$};

\draw (4, 0) -- (5,1);
\node at (6, 1)  {$\frac{p_a}{p_a+p_b+p_c+\ld{p_e}}$};

\draw (4, 0) -- (5,0);
\node at (6, 0)  {$\frac{p_b}{p_a+p_b+p_c+\ld{p_e}}$};

\draw (4, 0) -- (5,-1);
\node at (6, -1)  {$\frac{p_c}{p_a+p_b+p_c+\ld{p_e}}$};

\draw (4, 0) -- (5,-2);
\node at (6, -2)  {$\frac{\ld{p_e}}{p_a+p_b+p_c+\ld{p_e}}$};

\node at (4.5,0.7) {$a$};
\node at (4.5,0.2) {$b$};
\node at (4.5,-0.3) {$c$};
\node at (4.5,-0.8) {$e$};

\fill (7.4, -2) circle[radius=2pt];  % de finishes; paths abc
\draw (7.4, -2) -- (8.4,-1);
\draw (7.4, -2) -- (8.4,-2);
\draw (7.4, -2) -- (8.4,-3);

\node at (7.9,-1.2) {$p_a$};
\node at (7.9,-1.8) {$p_b$};
\node at (7.9,-2.3) {$p_c$};

\node at (9.3, -1) {$\frac{p_a}{p_a+p_b+c}$};
\node at (9.3, -2) {$\frac{p_b}{p_a+p_b+c}$};
\node at (9.3, -3) {$\frac{p_c}{p_a+p_b+c}$};

\fill (10.2, -1) circle[radius=2pt];  % dea finishes; paths bc
\fill (10.2, -2) circle[radius=2pt];  % deb finishes; paths ac
\fill (10.2, -3) circle[radius=2pt];  % dec finishes; paths ab

\draw (10.2, -1) -- (11.2,-0);
\node at (10.6,-0.4) {$b$};
\node at (10.6,-0.8) {$c$};

\node at (11.8, 0) {$\frac{\la{p_b}}{\la{p_b}+\la{p_c}}$};
\draw (10.2, -1) -- (11.2,-1);

\node at (11.8, -1) {$\frac{\la{p_c}}{\la{p_b}+\la{p_c}}$};

\draw (10.2, -2) -- (11.2,-1.6);
\node at (11.8, -1.6) {$\frac{\la{p_a}}{\la{p_a}+\la{p_c}}$};

\node at (10.6,-1.7) {$a$};
\node at (10.6,-2) {$c$};

\draw (10.2, -2) -- (11.2,-2.2);
\node at (11.8, -2.2) {$\frac{\la{p_c}}{\la{p_a}+\la{p_c}}$};

\draw (10.2, -3) -- (11.2,-2.9);
\node at (11.8, -2.9) {$\frac{\la{p_a}}{\la{p_a}+\la{p_b}}$};

\draw (10.2, -3) -- (11.2,-3.4);
\node at (11.8, -3.4) {$\frac{\la{p_b}}{\la{p_a}+\la{p_b}}$};

\node at (10.6,-2.8) {$a$};
\node at (10.6,-3.3) {$b$};

\fill (7.4, 1) circle[radius=2pt];  % da finishes; paths bce

\draw (7.4, 1) -- (8.4,2);
\draw (7.4, 1) -- (8.4,1);
\draw (7.4, 1) -- (8.4,0);

\node at (7.9,1.7) {$b$};
\node at (7.9,1.2) {$c$};
\node at (7.9,0.7) {$e$};

\node at (9.3, 2) {$\frac{\la{p_b}}{\la{p_b}+\la{p_c}+p_e}$};
\node at (9.3, 1) {$\frac{\la{p_c}}{\la{p_b}+\la{p_c}+p_e}$};
\node at (9.3, 0) {$\frac{e       }{\la{p_b}+\la{p_c}+p_e}$};

\fill (10.2, 1) circle[radius=2pt];  % da finishes; paths bce
\draw (10.2, 1) -- (11.2,2);
\node at (11.8, 2) {$\frac{\la{p_b}}{\la{p_b}+\ld{p_e}}$};
\draw (10.2, 1) -- (11.2,1);
\node at (11.8, 1) {$\frac{\ld{p_e}}{\la{p_b}+p_e}$};

\end{tikzpicture}
\caption{Partial probability\label{tikzabcde} tree for five competitors
  $a$-$e$ with mutually supporting subsets $\left\lbrace a,b,c\right\rbrace$
  [with support term $\lambda$] and $(de)$ [with support term $\mu$]}
\end{center}
\end{figure}

\begin{equation}\label{asuccb}
\frac{p_a}{p_a + p_b + p_c + p_d + p_e}\cdot
\frac{\lambda p_b}{\lambda p_b + \lambda p_c + p_d + p_e}\cdot
\frac{\lambda p_c}{\lambda p_c + p_d + p_e}\cdot
\frac{p_d}{p_d + p_e}
\end{equation}

In equation \ref{asuccb}, the first term is standard Plackett-Luce: at
this point, no competitor has finished and cheering effects are
absent.  The second term reflects the fact that competitors $b$ and
$c$ are supported by competitor $a$, who has by this point finished
the race and is supporting his teammates.

By contrast, the likelihood function for observation $d\succ a\succ
c\succ b\succ e$ would be

\begin{equation}
\frac{p_d}{p_a + p_b + p_c + p_d + p_e}\cdot
\frac{p_a}{p_a + p_b + p_c + \mu p_e}\cdot
\frac{\lambda p_c}{\lambda p_b + \lambda p_c + p_e}\cdot
\frac{\lambda p_b}{\lambda p_b + \mu p_e}
\end{equation}

where this likelihood function reflects the mutual support for
equivalence class $\left\lbrace d,e\right\rbrace$.  Note that the
final term reflects the fact that competitors $b$ and $e$ have their
support active when vying for fourth place, as members of both their
teams have finished at this point.

This probability model could easily be modified to account for
specific circumstances.  The cheering effect be asymmetrical (with $a$
helping $b$ but not vice-versa, for example).  The effect could
operate only on specific ordered pairs.  Or perhaps the effect might
have a finite lifetime: if $a$ places $n^\mathrm{th}$, then the
cheering effect is active only for competitors placing
$(n+r)^\mathrm{th}$ or above.
